\textsf{Are large language models superhuman chemists?}

### 2024-10-29 src/tex/appendix.tex
keep chembenchmini version

@@

\begin{figure}[htb]
    \centering
    \includegraphics[width=\textwidth]{figures/all_questions_models_completely_correct_radar_overall.pdf}
    \caption{\textbf{Performance of the models on the different topics of the \chembench corpus.} The radar plot shows the performance of the models on the different topics of the \chembench corpus. The performance is measured as the fraction of questions answered completely correctly by the models.
    A score of 1 (full coverage until the outer line of this plot) indicates that all questions were answered completely correctly, while a score of 0 indicates that none were answered completely correctly.
    }
    \label{fig:all_questions_models_completely_correct_radar_overall}
    \script{analyze_model_reports.py}
\end{figure}

\begin{figure}[htb]
    \centering
    \includegraphics{figures/performance_per_topic.pdf}
    \caption{\textbf{Fraction of correctly answered questions per data source.} The heatmap shows, in color, the fraction of questions answered correctly by different systems for some of our data sources. The performance is measured as the fraction of questions answered completely correctly by the models. A score of one (red) indicates that all questions were answered completely correctly, while a score of zero (blue) indicates that none of the questions were answered completely correctly.
        We see that the performance of the models varies significantly between the different data sources. For instance, it is interesting to observe that questions sourced based on textbooks seem easier for our leading models than for humans. However, this performance does not correlate with performance on other sources, e.g., semi-programmatically created tasks such as questions about the number of signals in an \gls{nmr} spectrum.
    }
    \label{fig:performance_per_topic}
    \script{analyze_performance_per_source.py}
\end{figure}

\begin{figure}[htb]
    \centering
    \includegraphics{figures/performance_per_topic_tiny.pdf}
    \caption{\textbf{Fraction of  correctly answered questions per data source on \chembenchmini.} The heatmap shows, in color, the fraction of questions answered completely correctly by different systems for some of our data sources. The performance is measured as the fraction of questions answered completely correctly by the models. A score of one (red) indicates that all questions were answered completely correctly, while a score of zero (blue) indicates that none were answered completely correctly.
        We see that the performance of the models varies significantly between the different data sources. For instance, it is interesting to observe that questions sourced based on textbooks seem easier for the leading models than for humans. However, this performance does not correlate with performance on other sources, e.g., semi-programmatically created tasks such as questions about the number of signals in an \gls{nmr} spectrum.
    }
    \label{fig:performance_per_topic_tiny}
    \script{analyze_performance_per_source.py}
\end{figure}

% In addition, we analyzed the performance on questions which requires calculation. For this, we manually labeled questions that require multiple calculation steps.
% We find that the ranking of models is different for questions with our without calculations (\Cref{fig:requires_cal}).

% \begin{figure}
%     \centering
%     \includegraphics{figures/model_overall_cal.pdf}
%     \caption{\textbf{Overall model performance for questions with and without calculation steps.} We find that the ranking of models changes if we evaluate them on questions that require and do not require calculations, respectively.}
%     \script{requires_calculation_plot.py}
%     \label{fig:requires_cal}
% \end{figure}

@@

\begin{figure}[htb]
    \centering
    \includegraphics{figures/human_score_distribution.pdf}
    \script{plot_human_score_distribution.py}
    \caption{\textbf{Distribution of human scores.} The histogram and kernel density estimates show the fraction of questions answered correctly by the human volunteers.
    Since the best possible score for each question is one and the worst possible score is zero, the values on this plot are between zero and one. A score of one would mean that a volunteer answered all questions correctly. A score of zero would mean that no question was answered correctly.}
    \label{fig:human_score_distribution}
\end{figure}

We also recorded the time humans took to answer the questions. This time is the time from the question being displayed to the human to the human submitting the answer.

@@

\clearpage
\subsection{Summary of trends and recommendations}

\oone consistently leads across both \ChemBench corpus and \chembenchmini.  \ClaudeThreeFiveSonnet, \GPTFourO, \LlamaThreeOneFourZeroFiveBInstruct form the next tier of strong performers. These models show robust performance across all skills and difficulty levels.

Larger models generally perform better (\LlamaThreeOneFourZeroFiveBInstruct > \LlamaThreeOneSeventyBInstruct > \LlamaThreeOneEightBInstruct). However, smaller but well-designed models can compete with larger models. For example, \GemmaTwoNineBIt has an overall accuracy of 0.48 on \ChemBench corpus, which is only 10\% less accurate than the much larger \LlamaThreeOneFourZeroFiveBInstruct model.
One can observe a clear progression of performance across model families (e.g., \ClaudeThreeFiveSonnet >  \ClaudeThree > \ClaudeTwo or \GPTFourO > \GPTFour > \GPTThreeFiveTurboZeroT). Newer versions consistently outperform their predecessors, indicating improvements in scientific understanding of these model families.
Scores on knowledge-intensive tasks are typically lower than those on calculation and reasoning-intensive questions. Performance on reasoning-intensive questions in \chembench tends to be the highest.

Based on these findings, some scope for improving these models could be to focus on enhancing knowledge-based training, possibly through improved pre-training on chemistry-specific texts or with chemistry knowledge-base integration. 
Models show the strongest performance in reasoning tasks on the \chembench corpus, which suggests that current training approaches are good at developing logical/reasoning capabilities for basic problem-solving of university-level exams. 
However, performance drops significantly for advanced tasks, suggesting that more advanced chemistry problems with complex, multi-step solutions should be included in training or finetuning.

\clearpage
\subsection{Leaderboard}
Our leaderboard is based on the tool chain developed for Matbench.\autocite{Dunn_2020}
Briefly, the \chembench pipeline produces standardized files in \texttt{json} format that contributors can add via pull requests to the \chembench repository.
The Markdown tables and interactive plots are automatically generated and updated on the \chembench website. The leaderboard is available at \url{https://lamalab-org.github.io/chem-bench/leaderboard/}.

>>>>

\begin{table}
    \caption{\textbf{Performance of the models on the \chembench corpus.} The table shows the fraction of questions answered correctly by the models for different skills and difficulty levels.}
    \resizebox{\textwidth}{!}{
    \variable{output/performance_table.tex}
    }
    \label{tab:performance_table}
\end{table}

\begin{table}
    \caption{\textbf{Performance of the models on the \enquote{tiny} subset of the \chembench corpus.} The table shows the fraction of questions answered correctly by the models for different skills and difficulty levels.}
    \resizebox{\textwidth}{!}{
    \variable{output/performance_table_human_subset.tex}
    }
    \label{tab:performance_table_human_subset}
\end{table}

\begin{figure}[htb]
    \centering
    \includegraphics[width=\textwidth]{figures/all_questions_models_completely_correct_radar_overall.pdf}
    \caption{\textbf{Performance of the models on the different topics of the \chembench corpus.} The radar plot shows the performance of the models on the different topics of the \chembench corpus. The performance is measured as the fraction of questions answered correctly by the models.
    A score of 1 (full coverage until the outer line of this plot) indicates that all questions were answered correctly, while a score of 0 indicates that none were answered correctly.
    }
    \label{fig:all_questions_models_completely_correct_radar_overall}
    \script{analyze_model_reports.py}
\end{figure}

\begin{figure}[htb]
    \centering
    \includegraphics{figures/performance_per_topic.pdf}
    \caption{\textbf{Fraction of correctly answered questions per data source on \chembenchmini.} The heatmap shows, in color, the fraction of questions answered correctly by different systems for some of our data sources. The performance is measured as the fraction of questions answered correctly by the models. A score of one (red) indicates that all questions were answered correctly, while a score of zero (blue) indicates that none of the questions were answered correctly.
        We see that the performance of the models varies significantly between the different data sources. For instance, it is interesting to observe that questions sourced based on textbooks seem easier for our leading models than for humans. However, this performance does not correlate with performance on other sources, e.g., semi-programmatically created tasks such as questions about the number of signals in an \gls{nmr} spectrum.
    }
    \label{fig:performance_per_topic}
    \script{analyze_performance_per_source.py}
\end{figure}

\Cref{fig:performance_corpus_and_tiny} shows the performance of the models on the \chembench corpus and the \chembenchmini subset. The relative performance difference between both is quite similar across most models. 
This makes \chembenchmini subset a reliable subset for human baseline comparison and particularly valuable for rapid prototyping and initial model assessment phases.

An interesting observation is the significant impact of chemical preference tasks on \GPTFour's scores. A detailed breakdown of overall accuracy into scores on different skills and difficulty levels is provided in \Cref{table:performance_table} and \Cref{tab:performance_table_human_subset} for \chembench corpus and the \chembenchmini subset, respectively.

\begin{figure}[htb]
    \centering
    \includegraphics{figures/performance_per_topic_tiny.pdf}
    \caption{\textbf{Fraction of completely correctly answered questions per data source on the \enquote{tiny} subset.} The heatmap shows, in color, the fraction of questions answered correctly by different systems for some of our data sources. The performance is measured as the fraction of questions answered correctly by the models. A score of one (red) indicates that all questions were answered correctly, while a score of zero (blue) indicates that none were answered correctly.
        We see that the performance of the models varies significantly between the different data sources. For instance, it is interesting to observe that questions sourced based on textbooks seem easier for the leading models than for humans. However, this performance does not correlate with performance on other sources, e.g., semi-programmatically created tasks such as questions about the number of signals in an \gls{nmr} spectrum.
    }
    \label{fig:performance_per_topic_tiny}
    \script{analyze_performance_per_source.py}
\end{figure}

\begin{figure}[htb]
    \centering
    \includegraphics{figures/corpus_human_comparison.pdf}
    \caption{\textbf{
        Performance of the models on the \chembench corpus and \chembenchmini subset.} The bar plot shows the fraction of questions that were answered completely correctly, highlighting the relative performance of different models on both the \chembench corpus and the \chembenchmini subset.
        We see that the model ranking remains fairly consistent across both sets}  
    \label{fig:performance_corpus_and_tiny}
    \script{corpus_humanset_performance.py}
\end{figure}

@@

@@

\begin{figure}[htb]
    \centering
    \includegraphics{figures/human_score_distribution.pdf}
    \script{plot_human_score_distribution.py}
    \caption{\textbf{Distribution of human scores.}The histogram and kernel density estimates show the fraction of questions answered correctly by the human volunteers.
    Since the best possible score for each question is one and the worst possible score is zero, the values on this plot are between zero and one. A score of one would mean that a volunteer answered all questions correctly. A score of zero would mean that no question was answered correctly. 
    We find that the scores for the questions that the human volunteers answered with tools are generally lower than the scores for the questions that the human volunteers answered without tools.}
    \label{fig:human_score_distribution}
\end{figure}

We also recorded the time humans took to answer the questions (\Cref{fig:human_timing}). This time is the time from the question being displayed to the human to the human submitting the answer.

@@

\clearpage
\subsection{Leaderboard}
\label{sec:leaderboard}
Our leaderboard is based on the tool chain developed for Matbench.\autocite{Dunn_2020}
Briefly, the \chembench pipeline produces standardized files in \texttt{json} format that contributors can add via pull requests to the \chembench repository.
The Markdown tables and interactive plots are automatically generated and updated on the \chembench website. The leaderboard is available at \url{https://lamalab-org.github.io/chem-bench/leaderboard/}.

### 2024-10-29 src/tex/ms.tex
keep chembenchmini version

Notably, we find that models are still limited in their ability to answer knowledge-intensive questions (\Cref{tab:performance_table_human_subset}), that is they did not memorize the relevant facts. Our results indicate that this is not a limitation that could be overcome by simple application of \gls{rag} systems such as \PaperQATwo. This is likely because the required knowledge cannot easily be accessed via papers (which is the only external knowledge \PaperQATwo has access to) but rather by lookup in specialized databases (e.g., PubChem, Gestis), which also the humans in our study used to answer such questions (\Cref{fig:tool_use}).
This indicates that there is still room for improvement of chemical \gls{llm} by training them on more specialized data sources or by integrating them with specialized databases.

In addition, our analysis shows that the performance of models is correlated with their size (see \Cref{fig:model_size_plot}). This is in line with observations in other domains but also indicates that chemical \gls{llm} could, to some extent, be further improved by scaling them up.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/all_questions_models_completely_correct_radar_human.pdf}
    \caption{\textbf{Performance of the models and humans on the different topics on \chembenchmini.} The radar plot shows the performance of the models and humans on the different topics of \chembenchmini. The performance is measured as the fraction of questions that were answered completely correctly by the models.
    The best score for every dimension is one (all questions answered correctly), and the worst is zero (no question answered correctly). A larger colored area indicates a better performance.
    This figure shows the performance on \chembenchmini. The performance of models on the entire corpus is shown in \Cref{fig:performance_per_topic}.
    }
    \label{fig:all_questions_models_completely_correct_radar_human}
    \script{analyze_model_reports.py}
\end{figure}

@@

\begin{table}[!h]
    \centering
    \caption{\textbf{Overview of sources of the curated questions}. The table provides an overview of the types of sources the questions have been curated from. Detailed sources are available in the source data on GitHub. Questions without source have been curated completely from scratch. Questions based on lecture notes or URL have been curated based on content presented in those resources. All questions have been rephrased, annotated, and reviewed before being added to the corpus. Note that the chemical preference questions are the main corpus of the semi-automatically generated group with \variable{output/question_count_per_dir/json_file_counts_preference.txt} questions of the total \variable{output/automatically_generated.txt}.}
    \variable{output/sources_table.tex}
    \label{tab:sources}
\end{table}

>>>>

Notably, we find that models are still limited in their ability to answer knowledge-intensive questions (\Cref{tab:performance_table_human_subset}), that is they did not memorize the relevant facts. Our results indicate that this is not a limitation that could be overcome by simple application of \gls{rag} systems such as \PaperQATwo. This is likely because the required knowledge cannot easily be accessed via papers (which is the only external knowledge \PaperQATwo has access to) but rather by lookup in specialized databases (e.g., PubChem, Gestis), which also the humans in our study used to answer such questions (\Cref{fig:tool_use}).
This indicates that there is still room for improvement of chemical \glspl{llm} by training them on more specialized data sources or by integrating them with specialized databases.

In addition, our analysis shows that the performance of models is correlated with their size (see \Cref{fig:model_size_plot}). This is in line with observations in other domains but also indicates that chemical \glspl{llm} could, to some extent, be further improved by scaling them up.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/all_questions_models_completely_correct_radar_human.pdf}
    \caption{\textbf{Performance of the models and humans on the different topics on \chembenchmini.} The radar plot shows the performance of the models and humans on the different topics of \chembenchmini. The performance is measured as the fraction of questions that were answered correctly by the models.
    The best score for every dimension is one (all questions answered correctly), and the worst is zero (no question answered correctly). A larger colored area indicates a better performance.
    This figure shows the performance on \chembenchmini. The performance of models on the entire corpus is shown in \Cref{fig:barplot_all_correct_all_questions}.
    }
    \label{fig:all_questions_models_completely_correct_radar_human}
    \script{analyze_model_reports.py}
\end{figure}

@@

\begin{table}[!h]
    \centering
    \caption{\textbf{Overview of sources of the curated questions}. The table provides an overview of the types of sources the questions have been curated from. Detailed sources are available in the source data on GitHub. Questions without source have been curated completely from scratch. Questions based on lecture notes or URL have been curated based on content presented in those resources. All questions have been rephrased, annotated, and reviewed before being added to the corpus.}
    \variable{output/sources_table.tex}
    \label{tab:sources}
\end{table}
